Countering language attrition with PanLex and the Web of Data

### 2014-03-08 panlex-rdf-paper/2012/MLOD_PanLex/paper_v5.tex
Update to the paper

The Semantic Web Initiative has led to a stack of technologies standardized by the
\emph{World Wide Web Consortium}\footnote{\url{http://www.w3.org/standards/semanticweb/}}
supporting a machine-readable and \mbox{-i}nter\-pretable Linked Data
network. This has led to a \emph{Web of Data}.
%Technologies that support this aim are already available and in use in the so
%called \emph{Web of Data}.
A growing community is leveraging Semantic Web technologies for linguistic knowledge, building a Web of Linguistic
Data, or a \emph{Linguistic Linked Open Data (LLOD) cloud}.

When linked open data are labeled in any language with lexemes in citation form (lemmas), the data are potentially linkable to, and through, PanLex. Here we describe how we connected PanLex to the Linked Data network.
In \autoref{sec:triplificate} we introduce the dataset, present a
PanLex RDF vocabulary, and explain how we transformed the one into the other and established conformance with additional data models.
\autoref{sec:linking} is about how we linked to other datasets of the LLOD cloud,
and \autoref{sec:publishing} is about the publication of the dataset.
Usage scenarios are given in \autoref{sec:usage}.
In \autoref{sec:related} we discuss related work, and, finally, in \autoref{sec:conclusion}, we summarize our recent and contemplated work.

\section{Triplification of the Raw Data}
\label{sec:triplificate}
Here we analyze the PanLex dataset; introduce our URI and vocabulary design, resembling PanLex's conceptual model;
%, which was developed independently from other ontology engineering approaches.
summarize how we classified PanLex's instance data with additional data models; and explain our transformation of the data to RDF.

@@

\subsection{Analysis of the Original Dataset}
\label{sec:analysis}
The PanLex \emph{database} is created by editors who consult sources\footnote{\url{http://panlex.org/tech/plrefs.shtml}}, such as mono- and multilingual dictionaries, glossaries, standards, and thesauri. The data include single- and multi-word expressions, meanings assigned to them, and related information. While copies of the sources would not be interoperable, PanLex data constitute editors' \emph{interpretations} of sources' \emph{assertions} that two or more \emph{expressions} share a \emph{meaning}\footnote{\url{http://panlex.org/tech/doc/design/panlex-db-design.pdf}}.
The most important entities and relations of PanLex's conceptual model are depicted in \autoref{fig:db-schema}.

@@

\subsection{Vocabulary Reuse}
\label{sec:vocabulary-reuse}
The PanLex vocabulary is based on PanLex's conceptual schema and
enables all of PanLex's data to be directly exposed as RDF.
%Apart from modeling a PanLex RDF vocabulary
Additionally, we also re-use existing vocuabularies, namely
the \emph{Lexicon Model for Ontologies} (lemon)~\cite{lemon2011} as well as the \emph{General Ontology for Linguistic Description} (GOLD)~\cite{farr2003}.
Since these models differ from the PanLex one to some extent, we follow an
incremental approach of aligning the PanLex data with them.
%accordingly. 
% one differ to some
%extend, not the whole schemas were implemented.
\autoref{tbl:sameclasses} shows PanLex classes with their current counterparts
in lemon and GOLD respectively.
The parts implemented in our RDF conversion are displayed in \autoref{fig:goldlemonimplemented}.
\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{images/pdf/gold_lemon_implemented.pdf}
  \caption{Parts of the GOLD (left) and lemon model (right) re-used in PanLex (URI prefixes are omitted for brevity)}
  \label{fig:goldlemonimplemented}
\end{figure}

\subsection{Transformation workflow}
\label{sec:conversion}
Since new sources are added to the PanLex database on almost a daily basis and because of its current size (\textasciitilde18 GB), the recurrent conversion of the database to capture changes in it is impractical.
%Using conventional hardware, a full conversion takes infeasibly long.
%This makes the testing and debugging of the modeled vocabularies and data very
% time-consuming.
As the PanLex data already reside in a relational database, the use of a virtual RDB2RDF\footnote{\url{http://www.w3.org/2001/sw/wiki/RDB2RDF}} mapping solution is a natural choice.
The \emph{Sparqlify system}\footnote{\url{https://github.com/AKSW/Sparqlify}} offers, besides an efficient query rewriting engine, also a very easy-to-use mapping language, called \emph{Sparqlification Mapping Language} (SML).
Essentially, these mappings consist of three clauses:
The \emph{From} clause specifies the logical SQL table (i.e. table, view, or
query) to be used in the SML view.
The \emph{With} clause binds a set of SPARQL variables to expressions that
yield RDF terms from relational columns.
Finally, the \emph{Construct} clause holds a set of triple patterns.
%clause making use of the defined SPARQL variable definitions.
\autoref{fig:ex:sparqlify-ml} shows an example of an SML view for the
languages in PanLex: From each row of the table \emph{i1} three resources are
created based on the \emph{iso3} column and 
bound to the variable names \emph{?lang}, \emph{?iso3} and \emph{?lexvo3}.
Resources for \emph{?lang} become typed as a \emph{Language} in the PanLex and the \emph{schema.org} namespace.
This view-based approach also demonstrates that changing the vocabulary or
adding support for new ones does not require an extract transform load (ETL) process,
and can therefore be done with little effort.

@@

\section{Linking}
\label{sec:linking}
The SML view in the previous section (\autoref{fig:ex:sparqlify-ml}) already established the interlinking of the PanLex languages with Lexvo.
In this section we outline the interlinking with DBpedia.
For DBpedia, we were interested in creating \emph{valid} and thus \emph{dereferenceable} links.
Therefore, we iterated the \emph{titles} datasets\footnote{\url{http://wiki.dbpedia.org/Downloads38}}, which map (non-localized) DBpedia URIs to their page titles in the respective language.
For each language version we normalized the labels by applying Unicode NFKD\footnote{\url{http://unicode.org/reports/tr15/}} normalization and removal of punctuation characters.
Each DBpedia resource was then mapped to the PanLex expression that was equal to the resource's normalized label in the respective language.
\autoref{fig:plx-dbp-link-counts} summarizes the number of links obtained.

@@

\section{Publishing}
\label{sec:publishing}
With our RDF conversion work, we complement existing
APIs\footnote{\url{http://panlex.org/try/}} with Linked Data, powered by
Pubby\footnote{\tiny{\url{http://wifo5-03.informatik.uni-mannheim.de/pubby/}}},
and two SPARQL
endpoints\footnote{\url{http://ld.panlex.org/vsparql}}
\footnote{\url{http://ld.panlex.org/sparql}}, ran by Sparqlify and Virtuoso,
respectively.
% and \url{http://ld.panlex.org/snorql}}.
An overview is shown in ~\autoref{fig:panlex-architecture}.
The SPARQL browser
\emph{SNORQL}\footnote{\url{https://github.com/kurtjx/SNORQL}} can be accessed
by replacing \emph{sparql} with \emph{snorql} in the respective links. Our SML
views and the interlinking code are hosted on GitHub\footnote{\url{https://github.com/AKSW/PanLex-2-RDF}}.
The created linksets are hosted in the PanLex database and are published together with the other data using the Sparqlify RDB2RDF tool.
Finally, we offer downloads tagged with timestamps of their creation\footnote{\url{http://ld.panlex.org/downloads/releases/}}.
\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{images/pdf/sparqlify_setup02.pdf}
\caption{PanLex architecture}
\label{fig:panlex-architecture}
\end{figure}

@@

We intend to address some limitations in the future. Attribution is currently limited to sources, obscuring the possible roles of multiple editors in iteratively interpreting sources and the subsidiary sources of data in sources that are themselves aggregations. Metadata attached to PanLex denotations are currently limited to arbitrary pairs of strings, but this sacrifices discovery possibilities when the metadata (as they often do) describe facts that can be expressed with PanLex expressions (e.g., that the expression is archaic). New collaborations between PanLex and related standards and technologies (e.g., language identification, language geolocation, lemmatization, transliteration, localization, etc.) are promising arenas for development. For example, the relations among PanLex's information sources, if treated as distinct datasets, could be modeled with the VoID vocabulary\footnote{\url{http://rdfs.org/ns/void}}.

\bibliographystyle{abbrv}
\bibliography{bibliography,../../bib/aksw}
\end{document}

>>>>

The Semantic Web Initiative has led to the development of standards and technologies
supporting a machine-readable and \mbox{-i}nter\-pretable Linked Data
network, known as the \emph{Web of Data}.\footnote{\url{http://www.w3.org/standards/semanticweb/}} From these efforts the \emph{Linked Open Data (LOD) cloud}\footnote{\url{http://lod-cloud.net/}} emerged.
%Technologies that support this aim are already available and in use in the so
%called \emph{Web of Data}.
A growing community is leveraging Semantic Web technologies for linguistic knowledge, building a \emph{Linguistic LOD (LLOD) cloud}.

Whenever there is Linked Data making use of descriptions, such as labels, in any language with lexemes in citation form (lemmas), the data are potentially linkable to, and through, PanLex. Here we describe how we connected PanLex to the Linked Data network.
In \autoref{sec:triplificate} we introduce the dataset, present a
PanLex RDF vocabulary, and explain how we transformed the one into the other and established conformance with additional data models.
\autoref{sec:linking} is about how we linked to other datasets of the LLOD cloud,
and \autoref{sec:publishing} is about the publication of the dataset.
Usage scenarios are given in \autoref{sec:usage}.
In \autoref{sec:related} we discuss related work, and, finally, in \autoref{sec:conclusion}, we summarize our recent and contemplated work.

\section{Triplification of the Raw Data}
\label{sec:triplificate}
In this section, we analyze the PanLex dataset; introduce our URI and vocabulary design, which resemble PanLex's conceptual model;
%, which was developed independently from other ontology engineering approaches.
summarize how we classified PanLex's instance data with additional data models; and explain our transformation of the data to RDF.

@@

\subsection{Analysis of the Original Dataset}
\label{sec:analysis}
The PanLex \emph{database} is created by editors who consult \emph{information sources}\footnote{\url{http://panlex.org/tech/plrefs.shtml}}, such as mono- and multilingual dictionaries, glossaries, standards, and thesauri. The data include single- and multi-word expressions, corresponding meanings assigned to them, and related information. While copies of the sources would not be interoperable, PanLex data constitute editors' \emph{interpretations} of sources' \emph{assertions} that two or more \emph{expressions} share a \emph{meaning}\footnote{\url{http://panlex.org/tech/doc/design/panlex-db-design.pdf}}.
The most important entities and relations of PanLex's conceptual model are depicted in \autoref{fig:db-schema}.

@@

\subsection{Vocabulary Reuse}
\label{sec:vocabulary-reuse}
The PanLex vocabulary is based on PanLex's conceptual schema and
enables all of PanLex's data to be directly exposed as RDF.
%Apart from modeling a PanLex RDF vocabulary
Additionally, we also re-use existing vocuabularies, namely
the \emph{Lexicon Model for Ontologies} (lemon)~\cite{lemon2011} as well as the \emph{General Ontology for Linguistic Description} (GOLD)~\cite{farr2003}.
Since these models differ from the PanLex one, we follow an
incremental approach of aligning the PanLex data with them.
%accordingly. 
% one differ to some
%extend, not the whole schemas were implemented.
\autoref{tbl:sameclasses} shows PanLex classes with their current counterparts
in lemon and GOLD respectively.
The parts implemented in our RDF conversion are displayed in \autoref{fig:goldlemonimplemented}.
\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{images/pdf/gold_lemon_implemented.pdf}
  \caption{Parts of the GOLD (left) and lemon model (right) re-used in PanLex (URI prefixes are omitted for brevity)}
  \label{fig:goldlemonimplemented}
\end{figure}

\subsection{Transformation workflow}
\label{sec:conversion}
Since new sources are added to the PanLex database on almost a daily basis and because of its current size (\textasciitilde18 GB), the recurrent conversion of the database to capture changes in it is impractical.
%Using conventional hardware, a full conversion takes infeasibly long.
%This makes the testing and debugging of the modeled vocabularies and data very
% time-consuming.
As the PanLex data already reside in a relational database, the use of a virtual RDB2RDF\footnote{\url{http://www.w3.org/2001/sw/wiki/RDB2RDF}} mapping solution is a natural choice.
The \emph{Sparqlify system}\footnote{\url{https://github.com/AKSW/Sparqlify}} offers, besides an efficient query rewriting engine, also a very easy-to-use mapping language, called \emph{Sparqlification Mapping Language} (SML).
Essentially, these mappings consist of three clauses:
The \emph{From} clause specifies the logical SQL table (i.e. table, view, or
query) to be used in the SML view.
The \emph{With} clause binds a set of SPARQL variables to expressions that
yield RDF terms from relational columns.
Finally, the \emph{Construct} clause holds a set of triple patterns.
%clause making use of the defined SPARQL variable definitions.
\autoref{fig:ex:sparqlify-ml} shows an example of an SML view for the
languages in PanLex: From each row of the table \emph{i1} three resources are
created based on the \emph{iso3} column and 
bound to the variable names \emph{?lang}, \emph{?iso3} and \emph{?lexvo3}.
Resources for \emph{?lang} become typed as a \emph{Language} in the PanLex and the \emph{schema.org} namespace.
This view-based approach makes it easy to perform future revisions of RDF mapping, such as
adding support for new vocabularies.
%, as this does not require an extract transform load (ETL) process.
%and can therefore be done with little effort.
%This view-based approach makes it easy to revise the RDF mapping, such as adding support for new vocabular

@@

\section{Linking}
\label{sec:linking}
The SML view in (\autoref{fig:ex:sparqlify-ml}) already establishes the interlinking of the PanLex languages with Lexvo.
In this section, we outline the interlinking with DBpedia, where we were interested in creating \emph{valid} and \emph{dereferenceable} links.
Therefore, we iterated the \emph{titles} datasets\footnote{\url{http://wiki.dbpedia.org/Downloads38}}, which map (non-localized) DBpedia URIs to their page titles in the respective language.
For each language version we normalized the labels by applying Unicode NFKD\footnote{\url{http://unicode.org/reports/tr15/}} normalization and removal of punctuation characters.
Each DBpedia resource was then mapped to the PanLex expression that was equal to the resource's normalized label in the respective language.
\autoref{fig:plx-dbp-link-counts} summarizes the number of links obtained.

@@

\section{Publishing}
\label{sec:publishing}
With our RDF conversion work, we complement existing
APIs\footnote{\url{http://panlex.org/try/}} with Linked Data, powered by
Pubby\footnote{\url{http://wifo5-03.informatik.uni-mannheim.de/pubby/}},
and two SPARQL
endpoints\footnote{\url{http://ld.panlex.org/vsparql}}
\footnote{\url{http://ld.panlex.org/sparql}}, ran by Sparqlify and Virtuoso.
% and \url{http://ld.panlex.org/snorql}}.
An overview is shown in ~\autoref{fig:panlex-architecture}.
The SPARQL browser
\emph{SNORQL}\footnote{\url{https://github.com/kurtjx/SNORQL}} can be accessed
by replacing \emph{sparql} with \emph{snorql} in the respective links. Our SML
views and the interlinking code are hosted on GitHub\footnote{\url{https://github.com/AKSW/PanLex-2-RDF}}.
The created linksets are hosted in the PanLex database and are published together with the other data using Sparqlify.
Finally, we offer downloads tagged with timestamps of their creation\footnote{\url{http://ld.panlex.org/downloads/releases/}}.
\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{images/pdf/sparqlify_setup02.pdf}
\caption{PanLex architecture}
\label{fig:panlex-architecture}
\end{figure}

@@

We intend to address some limitations in the future:
The relations among PanLex's information sources, if treated as distinct datasets, could be modeled with the VoID vocabulary\footnote{\url{http://rdfs.org/ns/void}}.
The (assertion) source entity should be refactored to model users and information sources as distinct entities.
%Attribution is currently limited to (assertion) sources,
% obscuring the possible roles of multiple editors in iteratively interpreting sources and the subsidiary sources of data in sources that are %themselves aggregations.
Metadata attached to PanLex denotations are currently limited to arbitrary pairs of strings,
but this sacrifices discovery possibilities when the metadata
%(as they often do)
describe facts that can again be expressed with PanLex expressions.
% (e.g., that the expression is archaic).
Finally, new collaborations between PanLex and related fields (e.g. as language identification, language geolocation, lemmatization, transliteration, localization, etc.) are promising arenas for development.

\bibliographystyle{abbrv}
\bibliography{bibliography,../../bib/aksw}

\end{document}
